---
title: "`webgem` Package"
format: html
---

::: {.package-hero}

`webgem` is an R package designed to help users summarize and visualize baseball performance data. The package includes functions for ranking player metrics, creating hitter profiles, and building radar-style visualizations that make player strengths and weaknesses easier to compare and understand.

*For deeper explanation and examples, visit the github repository* [{{< fa brands github >}}](https://github.com/ADC-405-S26/webgem)
:::

---

### Installation

*You can install the development version of `webgem` from GitHub with:*

```r
# install.packages("remotes")
# install.packages("devtools")
devtools::install_github("ADC-405-S26/webgem")
```

```r
library(webgem)
```

## Primary Functions

::: {.grid .function-grid}

::: {.g-col-4}
::: {.card .function-card}
### `metricRank()`

Ranks players based on selected performance metrics. This function helps compare athletes and identify which players stand out in specific areas.
:::
:::

::: {.g-col-4}
::: {.card .function-card}
### `hitterProfile()`

Creates a hitter summary profile that organizes important offensive information into a clean, readable format.
:::
:::

::: {.g-col-4}
::: {.card .function-card}
### `hitterRadar()`

Builds a radar-style chart to show a hitter’s strengths across multiple metrics. This makes it easier to compare performance categories visually.
:::
:::

:::

```{r}
#| echo: false
library(webgem)
```


## hitterRadar examples

**code portion**

::: {.panel-tabset}

## example 1

```{r}
#| echo: true
#| eval: false

library(webgem)

player_data <- webgem_data[webgem_data$player == "Drake Baldwin", ]

baldwin_profile <- hitterProfile(
  player = player_data$player,
  wrc_plus = player_data$wrc_plus,
  bb_pct = player_data$bb_pct,
  ops = player_data$ops,
  slg = player_data$slg,
  obp = player_data$obp
)

hitterRadar(baldwin_profile)


```


## example 2

```{r}
#| echo: true
#| eval: false

player_data <- webgem_data[webgem_data$player == "Dillon Dingler", ]

dingler_profile <- hitterProfile(
  player = player_data$player,
  slg = player_data$slg,
  ops = player_data$ops,
  avg = player_data$avg,
  iso = player_data$iso,
  obp = player_data$obp,
  war = player_data$war
)

hitterRadar(dingler_profile)

```

## example 3

```{r}
#| echo: true
#| eval: false

player_data <- webgem_data[webgem_data$player == "Rafael Devers", ]

devers_profile <- hitterProfile(
  player = player_data$player,
  ops = player_data$ops,
  avg = player_data$avg,
  woba = player_data$woba,
  barrel_pct = player_data$iso,
  iso = player_data$iso,
  wrc_plus = player_data$wrc_plus
)

hitterRadar(devers_profile)

```

## example 4

```{r}
#| echo: true
#| eval: false

player_data <- webgem_data[webgem_data$player == "Yordan Alvares", ]

alvares_profile <- hitterProfile(
  player = player_data$player,
  avg = player_data$avg,
  bb_pct = player_data$bb_pct,
  hard_hit_pct = player_data$hard_hit_pct,
  obp = player_data$obp,
  wrc_plus = player_data$wrc_plus
)

hitterRadar(alvares_profile)

```

:::

**plot portion**

*code ex. number correlates to the numbered plot*

::: {.panel-tabset}

## example 1

```{r}
#| echo: false

player_data <- webgem_data[webgem_data$player == "Drake Baldwin", ]

baldwin_profile <- hitterProfile(
  player = player_data$player,
  wrc_plus = player_data$wrc_plus,
  bb_pct = player_data$bb_pct,
  ops = player_data$ops,
  slg = player_data$slg,
  obp = player_data$obp
)

hitterRadar(baldwin_profile)

```

## example 2

```{r}
#| echo: false

player_data <- webgem_data[webgem_data$player == "Dillon Dingler", ]

dingler_profile <- hitterProfile(
  player = player_data$player,
  slg = player_data$slg,
  ops = player_data$ops,
  avg = player_data$avg,
  iso = player_data$iso,
  obp = player_data$obp,
  war = player_data$war
)

hitterRadar(dingler_profile)

```

## example 3

```{r}
#| echo: false

player_data <- webgem_data[webgem_data$player == "Rafael Devers", ]

devers_profile <- hitterProfile(
  player = player_data$player,
  ops = player_data$ops,
  avg = player_data$avg,
  woba = player_data$woba,
  barrel_pct = player_data$iso,
  iso = player_data$iso,
  wrc_plus = player_data$wrc_plus
)

hitterRadar(devers_profile)

```

## example 4

```{r}
#| echo: false

player_data <- webgem_data[webgem_data$player == "Yordan Alvares", ]

alvares_profile <- hitterProfile(
  player = player_data$player,
  avg = player_data$avg,
  bb_pct = player_data$bb_pct,
  hard_hit_pct = player_data$hard_hit_pct,
  obp = player_data$obp,
  wrc_plus = player_data$wrc_plus
)

hitterRadar(alvares_profile)

```

:::
