---
title: "Text Analysis Project"
page-layout: full
format: html
---

<i class="bi bi-github"></i> <a
href="https://github.com/ADC-405-S26/Project_3_TextAnalysis_Team_7" target="_blank">GitHub Repo</a>

Note: This was a joint project done with Will Zatkowski

## Introduction & Background

Religious texts have shaped history not only through what they say, but also through how they are written. The four Gospels in the Bible are often read together because they each tell the story of Jesus from four different perspectives. While they overlap in many places, they also differ in structure, emphasis, and wording. These differences make the Gospels a useful case for text analysis, where we can compare not just what each Gospel says, but how each one says it. In this blog post, we will use computational tools to explore whether the Gospels’ language patterns reflect their historical and literary relationships. To be more specific, our primary goal is to identify which words and themes are most distinctive from one Gospel to the next, and understand the linguistic differences tied to the unique writing styles of each author.


For a deeper dive into the historical background, check out our [PDF document!](https://drive.google.com/file/d/1N7kFC9q3tKnOqyLSO3LXhZtbjc_K8R5H/view?usp=sharing)

## Data

The data used in this project contains the King James version of the four Gospels: Matthew, Mark, Luke, and John, from Gutenberg metadata. Each observation represents a portion of Gospel text, organized by Gospel and verse-level information. The raw version of our initial dataset includes the Gutenberg-assigned ID for each gospel, the free-form text, the title, and the Gospel name.
```{r}
#| include: false

library(tidyverse)
library(scales)
library(tidytext)
library(textdata)
library(gutenbergr)
library(stylo)

# install.packages("tidyverse")
# install.packages("scales")
# install.packages("tidytext")
# install.packages("textdata")
# install.packages("gutenbergr")
# install.packages("stylo")

```

```{r}
#| include: false
#| fig-align: "center"

data("gutenberg_metadata")

kingJamesBible_all <- gutenberg_metadata |>
  filter(str_detect(title, "Bible"))
# accessing data from the gutenberg library, checking for version of bible
# The King James Bible

Matthew <- gutenberg_download(8040, meta_fields = "title")
Mark <- gutenberg_download(8041, meta_fields = "title")
Luke <- gutenberg_download(8042, meta_fields = "title")
John <- gutenberg_download(8043, meta_fields = "title")
# getting the exact ID for the entry

```


```{r}
#| include: false

# downloading files into one dataset

gospel_ids <- tibble(
  gutenberg_id = c(8040, 8041, 8042, 8043),
  gospel = c("Matthew", "Mark", "Luke", "John")
)

gospels_raw <- gutenberg_download(
  gospel_ids$gutenberg_id,
  meta_fields = "title"
) |>
  left_join(gospel_ids, by = "gutenberg_id")

glimpse(gospels_raw)

```

```{r}
#| echo: false

gospels_clean <- gospels_raw |>
  filter(text != "") |>   # removing any rows where the text is now blank.

  # removing lines that begin with "book" followed by a space.
  # str_detect() checks whether a text pattern is present.
  filter(!str_detect(text, regex("^book\\s+", ignore_case = TRUE))) |>
  mutate(
    # str_replace_all() replaces every match of a pattern.
    # \\d+ - "one or more digits."
    # this removes chapter numbers, verse numbers, or labels
    text = str_replace_all(text, "\\d+", " "),
    
    # removing leftover colons from the beginning of the line
    # \\s* - "zero or more spaces"
    # : - "match a colon"
    # :? - "optionally matches one more colon"
    # the "" at the end deletes the matched punctuation
    text = str_replace(text, "^\\s*:\\s*:?\\s*", ""), 
    
    text = str_squish(text) # str_squish() removes extra spaces from each line of text.
  )
```

::: {.panel-tabset}

## Uncleaned Data

```{r}
#| echo: false

knitr::kable(
  head(gospels_raw),
  caption = "First few rows of the uncleaned Gospel text data"
)
```

## Cleaning Process

```{r}
#| echo: true
#| eval: false

gospels_clean <- gospels_raw |>
  filter(text != "") |>   # removing any rows where the text is now blank.

  # removing lines that begin with "book" followed by a space.
  # str_detect() checks whether a text pattern is present.
  filter(!str_detect(text, regex("^book\\s+", ignore_case = TRUE))) |>
  mutate(
    # str_replace_all() replaces every match of a pattern.
    # \\d+ - "one or more digits."
    # this removes chapter numbers, verse numbers, or labels
    text = str_replace_all(text, "\\d+", " "),
    
    # removing leftover colons from the beginning of the line
    # \\s* - "zero or more spaces"
    # : - "match a colon"
    # :? - "optionally matches one more colon"
    # the "" at the end deletes the matched punctuation
    text = str_replace(text, "^\\s*:\\s*:?\\s*", ""), 
    
    text = str_squish(text) # str_squish() removes extra spaces from each line of text.
  )
```

## Cleaned Data

```{r}
#| echo: false

knitr::kable(
  head(gospels_clean),
  caption = "First few rows of the cleaned Gospel text data"
)
```

:::

## Exploring Distinctive Thematic Language in the four Gospels

```{r}
#| include: false

library(tidyverse)
library(tidytext)
library(quanteda)
library(knitr)
library(wesanderson)

# manually creating themes for different words we want to focus on for tf-idf analysis
# also, because many of these words were not assigned a score through sentiment analysis,
# manually assigning range from -2 to 2
focus_words <- tribble(
  ~word, ~connotation, ~theme,
  
  # theme forgiveness
  "forgive", 2, "forgiveness",
  "forgiven", 2, "forgiveness",
  "repent", 1, "forgiveness",
  "repentance", 1, "forgiveness",
  
  # theme love
  "love", 2, "love",
  "loved", 2, "love",
  "loveth", 2, "love",
  "mercy", 2, "love",
  "merciful", 2, "love",
  "compassion", 2, "love",
  "pity", 1, "love",
  "grace", 2, "love",
  "way", 1, "love",
  "life", 0, "love",
  "light", 2, "love",
  
  # theme miracles
  "miracle", 2, "miracles",
  "miracles", 2, "miracles",
  "sign", 1, "miracles",
  "signs", 1, "miracles",
  "heal", 2, "miracles",
  "healed", 2, "miracles",
  "healing", 2, "miracles",
  "sick", 0, "miracles",
  "blind", -1, "miracles",
  "lame", -1, "miracles",
  "leper", 0, "miracles",
  "mighty works", 2, "miracles",
  "wonders", 2, "miracles",
  
  # theme evil
  "devils", -2, "evil",
  "devil", -2, "evil",
  "demon", -2, "evil",
  "demons", -2, "evil",
  "unclean", -2, "evil",
  "darkness", -2, "evil",
  "tempted", -2, "evil",
  "betray", -2, "evil",
  "betrayed", -2, "evil",
  "accuse", -2, "evil",
  "accused", -2, "evil",
  "enemy", -2, "evil",
  "enemies", -2, "evil",
  "pharisees", 0, "evil",
  "scribes", 0, "evil",
  "sin", -2, "evil",
  "sins", -2, "evil",
  "sinner", -2, "evil",
  "sinners", -2, "evil",
  "judgment", -2, "evil",
  "condemn", -2, "evil",
  "condemned", -2, "evil",
  
  # theme faith
  "faith", 2, "faith",
  "believe", 2, "faith",
  "believed", 2, "faith",
  "believeth", 2, "faith",
  "belief", 2, "faith",
  "witness", 1, "faith",
  "testimony", 1, "faith",
  "testify", 1, "faith",
  "truth", 2, "faith",
  "abide", 1, "faith",
  "dwelt", 1, "faith",
  "spirit", 0, "faith",
  "risen", 1, "faith",
  "rise", 1, "faith",
  "resurrection", 2, "faith",
  "tomb", 0, "faith",
  "reborn", 1, "faith",
  "born again", 1, "faith",
  "glory", 2, "faith",
  "jesus", 1, "faith",
  "son of god", 1, "faith",
  
  # theme death
  "death", -2, "death",
  "die", -2, "death",
  "died", -2, "death",
  "kill", -2, "death",
  "killed", -2, "death",
  "crucify", -2, "death",
  "crucified", -2, "death",
  "cross", -2, "death",
  "buried", -1, "death",
  "blood", -1, "death",
  

) |>
  distinct(word, .keep_all = TRUE)
```


```{r}
#| include: false

# cleaning focus_word df
focus_words_clean <- focus_words |>
  mutate(
    
    # str_replace_all() here replaces spaces with underscores for phrase matching
    # i.e.: "son of god" --> "son_of_god"
    feature = str_replace_all(word, "\\s+", "_")) |>
  
  # keeping only one row per feature
  distinct(feature, .keep_all = TRUE)


# pulling out only the multi-word phrases from the focus words
# str_detect here checks whether the word contains a space
# pull(word) then turns that column into a vector
multiword_terms <- focus_words_clean |>
  filter(str_detect(word, "\\s+")) |>
  pull(word)
```

```{r}
#| include: false

# collapsing the verses into one document per Gospel

gospel_docs <- gospels_clean |>
  group_by(gospel) |>
  summarize(
    
    # str_c() combines all verse text from each Gospel into one long string
    text = str_c(text, collapse = " "),
    .groups = "drop")

# creating quanteda corpus object for storing the texts
gospel_corpus <- corpus(
  gospel_docs,
  text_field = "text")

docnames(gospel_corpus) <- gospel_docs$gospel

# tokenizing the text
gospel_tokens <- gospel_corpus |>
  tokens(
    remove_punct = TRUE, # removes punctuation
    remove_numbers = TRUE,  # removes numbers
    remove_symbols = TRUE   # removes symbols
  )

# combining the multi-word phrases into single tokens
# this is necessary so multiple word phrases aren't counted separately
# tokens_compound() keeps selected phrases together
# clarifying the 'concatenator' connecting these phrases
if (length(multiword_terms) > 0) {
  gospel_tokens <- gospel_tokens |>
    tokens_compound(
      pattern = phrase(multiword_terms),
      concatenator = "_")}

# creating document-feature matrix, necessary for using quanteda
# this tell us how many times each feature appears in each Gospel
# dfm() creates the matrix
gospel_dfm <- dfm(gospel_tokens)

# now filtering out only the previously determined focus words
# dfm_select() filters the matrix keeping only this
focus_dfm <- dfm_select(
  gospel_dfm,
  pattern = focus_words_clean$feature,
  selection = "keep",
  valuetype = "fixed")

# dfm_tfidf() calculates the raw tokens into tf-idf scores
# scheme_tf = "prop" uses the proportional word frequency (accounts for different Gospel lengths)
focus_tfidf <- dfm_tfidf(
  focus_dfm,
  scheme_tf = "prop")

```


```{r}
#| include: false

# converting the TF-IDF matrix into long format for plotting purposes
focus_tfidf_tidy <- as.data.frame(
  as.matrix(focus_tfidf),
  # check.names = FALSE keeps R from changing feature names
  check.names = FALSE) |>
  
  # rownames_to_column() turns Gospel names into a real column
  rownames_to_column("document") |>
  
  # converting from wide to long format 
  pivot_longer(
    cols = -document,
    names_to = "feature",
    values_to = "tfidf") |>
  
  # rename() changes document to gospel
  rename(gospel = document) |>
  
  # joining the tf-idf results back to focus word df
  left_join(
    focus_words_clean,
    by = "feature") |>
  
  # calculating the weighted contribution score of each feature
  # this score represents both how important the word is to the specific Gospel,
  # and how strongly positive or negative (sentiment, not numeric) it is
  mutate(
    word_contribution = tfidf * connotation)
```

```{r}
#| include: false

# checking to see tf-idf scores against and calculated metric
focus_tfidf_tidy |>
  filter(tfidf > 0) |>
  arrange(desc(tfidf)) |>
  head(25)

# shows which words were not included in Gospels form selected focus words
setdiff(focus_words_clean$feature, featnames(focus_dfm))

```
 

```{r}
#| include: false

# because some themes contain more words than others, counting the number of terms 
# in each theme to scale them and create one normalized score
theme_sizes <- focus_words_clean |>
  count(theme, name = "number_of_terms")

# creating one score for each Gospel and theme combination
theme_scores <- focus_tfidf_tidy |>
  group_by(gospel, theme) |>
  summarize(
    
    # total_tfidf shows how distinctive the theme is,
    # and ignores whether the connotation is positive or negative
    total_tfidf = sum(tfidf),
    
    # raw_theme_score adds up tf-idf × connotation for all words in the theme
    # positive theme words increase score, vice-versa
    raw_theme_score = sum(word_contribution),
    .groups = "drop") |>
  
   # Add the number of focus terms in each them
  left_join(theme_sizes, by = "theme") |>
  
   # now normalizing the score by dividing by the number of terms in the theme
   # this will be used in heatmap
  mutate(
    normalized_theme_score = raw_theme_score / number_of_terms)
   

head(theme_scores)

```

We first wanted to compare the Gospels through the specific types of language they emphasize. To do so, we split core words and phrases throughout the bible into 6 themes for later sentiment analysis: forgiveness, love, faith, miracles, evil, and death. Focusing on a smaller subset of words forces the analysis to center more directly on the theological and emotional language that shapes each Gospel. However, using sentiment analysis with religious phrases and wording can become quite buggy. Therefore, we had to manually create a 'lexicon' that not only categorized the custom set of words into these themes, but also gave a connotation score from -2 to 2: negative values represent darker or conflict-oriented language and positive values represent redemptive or hopeful language.

Since TF-IDF only shows which words are most distinctive in each Gospel, we combined it with our connotation scores to make the measure more interpretive. In other words, a weighted connotation score was created by multiplying each word’s TF-IDF value by its manually assigned connotation score.

```{r}
#| echo: false

# THIS IS NOT NEEDED FOR ACTUAL HEATMAP

# small table for demonstration in final pdf
focus_tfidf_tidy |>
  filter(tfidf > 0) |>
  arrange(desc(tfidf)) |>
  select(
    Gospel = gospel,
    Word = word,
    Theme = theme,
    Connotation = connotation,
    `TF-IDF` = tfidf,
    `Weighted contribution` = word_contribution) |>
  
  # showing only 4 decimals
  mutate(
    `TF-IDF` = round(`TF-IDF`, 4),
    `Weighted contribution` = round(`Weighted contribution`, 4)) |>
  head(10) |>
  
  # knitr::kable() is better for formatting the table so it looks cleaner in final render
  knitr::kable(
    caption = "Sample of focus words after converting TF-IDF output into tidy format")

```

Now we visualized these scores using a heatmap, with the x-axis comparing Matthew, Mark, Luke, and John, and the y-axis listing the six themes. Each tile represents one Gospel-theme combination, with the color and number showing whether that theme is more positively or negatively weighted in that Gospel. Strong positive values indicate that a theme is being shaped by words that are both distinctive to that Gospel and positively coded. Strong negative values indicate that a theme is being shaped by words that are both distinctive to that Gospel and negatively coded. These values are the normalized theme scores multiplied by 1000, which was done only to make the very small TF-IDF-based values easier to read on the plot.

---

```{r, fig.width=9, fig.height=6}
#| echo: false
#| message: false
#| warning: false
#| results: hide
#| fig-align: "center"

# testing palettes
rocket_colors <- wes_palette("BottleRocket2", n = 4)
cavalcanti_colors <- wes_palette("Cavalcanti1", n = 4)

# adjusting numeric score for readability purposes
theme_scores_display <- theme_scores |>
  mutate(
    display_score = normalized_theme_score * 1000)

# making the color scale for heatmap (balanced around 0)
max_abs_score <- max(abs(theme_scores_display$display_score), na.rm = TRUE)

ggplot(theme_scores_display,
  aes(gospel, theme, fill = display_score)) +
  
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = number(display_score, accuracy = 0.1)),
    size = 3.5) +
  
  scale_fill_gradient2(
    low = "brown1",
    mid = "white",
    high = "cyan3",
    midpoint = 0,
    limits = c(-max_abs_score, max_abs_score),
    labels = number_format(accuracy = 0.1),
    name = "Score\n×1000") +
  
  labs(
    title = "Thematic TF-IDF x Connotation Scores Across the Four Gospels",
    subtitle = "Scores combine focus-word distinctiveness with positive or negative connotation ratings",
    x = "Gospel",
    y = "Theme") +
  
  theme_void() +
  theme(
    plot.background = element_rect(color = "#2c3e50", linewidth = 0.4),
    plot.margin = margin(t = 8, r = 8, b = 8, l = 8),
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 11, face = "bold"))

```

Luke, Mark, and Matthew all demonstrate strong similarities in the forgiveness, miracles, and faith themes. Forgiveness appears to be heavily significant across these three Gospels, with Luke having the highest weighted connotation score at 2.3, followed by Mark at 1.8 and Matthew at 1.7. In contrast, the textual data in John results in a value of 0...

Check out [our pdf](https://drive.google.com/file/d/1N7kFC9q3tKnOqyLSO3LXhZtbjc_K8R5H/view?usp=sharing) for a deeper analyis.

